Predicting the effect of candidate interventions for mitigating disease spread in asymmetric-mixing metapopulations

### 2024-07-24 ms.tex
connecting computational analysis subsec with analysis sketch

To equitably prepare for pandemics in our uncertain world we need to
\autocite{Rodo2021,Adashi2022}. The global COVID-19 pandemic highlighted
the current lack of outcome equality, where a country's inequality
(measured by the Gini index) predicted higher burden of COVID-19
infections \autocite{Su2022}. Here we contribute to this program of more
equitable pandemic preparedness by developing a compartmental
behavioral-epidemiological metapopulation model of COVID-19 transmission
between a Brazilian city, Dourados, in Mato Grosso do Sul, and the
neighboring Dourados Indigenous Reserve \autocite{DeOliveira2023}.
COVID-19 burden was felt disproportionately by Indigenous peoples in
Brazil \autocite{Simionatto2020}, but may have been worse in the
Dourados Reserve due to its immediate proximity to the Dourados. Nearly
all residents of the reserve live in domiciles in one of two villages in
the Reserve, Jaguapiru and Bororó. Jaguapiru is closer overall to the
city of Dourados, sharing a much longer border with the city limits than
Bororó (Figure 1). Indigenous peoples on the reserve are predominantly
one of two ethnicities, Guarani and Terena. Anecdotally, Terena are
known for being better educated with somewhat more robust physicality
and, perhaps, better resistance to disease. They may travel to the city
more frequently for trade and other purposes.

De Oliveira and co-authors hypothesized that COVID-19 variant diffusion
from Dourados City into the Dourados Reserve is driven by young
Indigenous men traveling between the reserve and the city of Dourados
for economic reasons. This is a reasonable hypothesis given available
data.

\mt{IF TERENA ARE MINORITY, HAVE HIGH MOBILITY TO CITY, AND $c_{GT} > c_{TG}$
THERE COULD BE A MINORITY INCUBATOR-TYPE EFFECT~\autocite{Turner2023}}

@@

Interventions center around reducing the mobility of Indigenous people between
Dourados city and Reserve; the total change in mobility, denoted $\Delta
\gamma$, is the total reduction of the parameter $\gamma_g$ across all 
groups, i.e., 
\begin{equation}
  \Delta \gamma = \sum_g \Delta \gamma_g.
\end{equation}
\noindent
Different interventions allocate $\Delta \gamma$ to different degrees. 
One strategy we test is the \emph{Equal} strategy, where all $\gamma_g$ are
equal. Another named strategy is the \emph{80-20} strategy, where the
reduction in mobility in group $g$ is 80\% of the total, and reduction in
group $g'$ mobility is 20\% of the total, i.e., $\Delta\gamma_g / \Delta\gamma =
0.8$ and $\Delta\gamma_{g'} / \Delta\gamma = 0.2$ To create alternative
strategies, both the total reduction and the ratio of group-level reduction to 
total reduction can be changed. We finally evaluate the hypothesis that
interventions that may be effective for strains with no re-infection (like
B.1.1) will not be equally effective for strains with re-infection (like Zeta),
and vice-versa. This will highlight the importance of accounting for social and
epidemiological context, alike, in pandemic preparedness via this program of
computational evaluation of intervention strategies.

>>>>

To equitably prepare for pandemics in our uncertain world we
need to predict the effect of candidate interventions to mitigate novel spillover
disease and novel variants of existing disease
\autocite{Rodo2021,Adashi2022}. The global COVID-19 pandemic highlighted
the current lack of outcome equality, where a country's inequality
(measured by the Gini index) predicted higher burden of COVID-19
infections \autocite{Su2022}. Here we contribute to this program of more
equitable pandemic preparedness by developing a compartmental
behavioral-epidemiological metapopulation model with an immediate application:
disease transmission between a Brazilian city, Dourados, in the state of 
Mato Grosso do Sul, and the
neighboring Dourados Indigenous Reserve \autocite{DeOliveira2023}.
COVID-19 burden was felt disproportionately by Indigenous peoples throughout
Brazil \autocite{Simionatto2020}, but the burden may have been even more 
inequitable in the
Dourados Reserve due to its immediate proximity to the city of Dourados. Nearly
all residents of the reserve live in domiciles in one of two villages in
the Reserve, Jaguapiru and Bororó. Jaguapiru is closer overall to the
city of Dourados, sharing a much longer border with the city limits than
Bororó (Figure \ref{fig:Map}). Indigenous peoples on the reserve are predominantly
one of two ethnicities, Guarani and Terena. 
% Anecdotally, Terena are
% known for being better educated with somewhat more robust physicality
% and, perhaps, better resistance to disease. 
% Anecdotally Terena may travel to the city more frequently for trade and other purposes,
% however De Olivera and colleagues 
De Oliveira and co-authors hypothesized that COVID-19 variant diffusion
from Dourados City into the Dourados Reserve is driven by young
Indigenous men traveling between the reserve and the city of Dourados
for economic reasons. This is a reasonable starting hypothesis given available
data, but it points to potentially richer, more explanatory mechanisms and 
a more detailed set of hypotheses. Understanding transmission in the Reserve, and
evaluating possible interventions based on this understanding, could mitigate
the next pandemic that affects the Dourados Reserve, and lessons learned here could
transfer to pandemic preparedness planning in other contexts.

@@

Interventions center around reducing the mobility of Indigenous people between
Dourados city and Reserve; the total change in mobility, denoted $\Delta
\gamma$, is the total reduction of the parameter $\gamma_g$ across all 
groups, i.e., 
\begin{equation}
  \Delta \gamma = \sum_g \Delta \gamma_g.
\end{equation}
\noindent
Different interventions allocate $\Delta \gamma$ to different degrees. 
One strategy we test is the \emph{Equal} strategy, where all $\gamma_g$ are
equal. Another named strategy is the \emph{80-20} strategy, where the
reduction in mobility in group $g$ is 80\% of the total, and reduction in
group $g'$ mobility is 20\% of the total, i.e., $\Delta\gamma_g / \Delta\gamma =
0.8$ and $\Delta\gamma_{g'} / \Delta\gamma = 0.2$. To create alternative
strategies, both the total reduction and the ratio of group-level reduction to 
total reduction can be changed. We finally evaluate the hypothesis that
interventions that may be effective for strains with no re-infection (like
B.1.1) will not be equally effective for strains with re-infection (like Zeta),
and vice-versa. This will highlight the importance of accounting for social and
epidemiological context, alike, in pandemic preparedness via this program of
computational evaluation of intervention strategies.
